---
title: "IDL Research Project"
author: "Lamar Foster"
date: "10/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
```
The Department of Housing and Urban Development (HUD) considers people spending more than 30% of their income on housing as housing burdened. Housing costs include rent, mortgage payement, property taxes, utilities, and housing insurance (Lin, 2018). There has recently been a spotlight on how energy prices, i.e. utilities, contribute to housing burden (Buylova, 2020). Energy bills of more than 6-10% of income is considered unaffordable (Hernandez & Bird, 2010; Heindl, 2015; Kontokosta et al., 2019). Energy upgrades can  help alleviate housing burden by saving a low-income multifamily household up to $1,500 per year per household (Kontokosta et al., 2019).

Energy burden is a function of both income and energy consumption. Drehobl and Ross (2016) found that an average low-income household energy burden was 7.2%, which is higher than an average higher income household energy burden of 2.3%. Findings from an evaluation of a federal energy assistance program determiined that low-income households represent 92% of all households with energy burden of 10% (APPRISE, 2005). Federal Low-Income Home Energy Assistance Program (LIHEAP) designs its funding distribution mechanism based on energy burden levels. Income is used as the main qualifier for this program.

Buyolva (2020) used two datasets: a household energy audit survey - the residential building stock assessment (RBSA) completed by the Northwest Energy Efficiency Association (NEEA) in 2017 and the American Community Survey (ACS) 2012-2017 5-year estimates data. 

Suggested Readings: 

[Buylova, A. (2020). Spotlight on energy efficiency in Oregon: Investigating dynamics between energy use and socio-demographic characteristics in spatial modeling of residential energy consumption. Energy Policy, 140, 111439.](https://www.sciencedirect.com/science/article/pii/S0301421520301920)

[Kontokosta, C. E., Reina, V. J., & Bonczak, B. (2020). Energy cost burdens for low-income and minority households: Evidence from energy benchmarking and audit data in five US cities. Journal of the American Planning Association, 86(1), 89-105.](https://www.tandfonline.com/doi/full/10.1080/01944363.2019.1647446)

---
This data is from the Residential Building Stock Assessment [(RBSA)](https://neea.org/resources/rbsa-ii-combined-database) conducted in 2016-17. It includes a representative sample of single-family, Multi-family and manufactured homes gathered across the Northwest region. Whenever possible, data was collected in a similar manner as the 2011 – 2012 assessment to ensure continuity and comparability between the studies. Supporting documentation of the RBSA data can be found [here](https://neea.org/data/residential-building-stock-assessment). 

```{r, message=FALSE, warning=FALSE}
ols_df <- read_csv("RBSA-II-Combined-Database/One Line Summary.csv")
dictionary_df <- read_csv("RBSA-II-Combined-Database/DataDictionary.csv")
demog <- read_csv("RBSA-II-Combined-Database/SiteInterview_Demographics.csv")
```

