---
title: "IDL Research Project"
author: "Lamar Foster"
date: "10/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
```

This data is from the Residential Building Stock Assessment [https://neea.org/resources/rbsa-ii-combined-database](RBSA) conducted in 2016-17. It includes a representative sample of single-family, Multi-family and manufactured homes gathered across the Northwest region. Whenever possible, data was collected in a similar manner as the 2011 – 2012 assessment to ensure continuity and comparability between the studies. Supporting documentation of the RBSA data can be found [https://neea.org/data/residential-building-stock-assessment](here). 


```{r}
ols_df <- read_csv("~/Desktop/DS Projects/RBSA-II-Combined-Database/One Line Summary.csv")
dictionary_df <- read_csv("~/Desktop/DS Projects/RBSA-II-Combined-Database/DataDictionary.csv")
demog <- read_csv("~/Desktop/DS Projects/RBSA-II-Combined-Database/SiteInterview_Demographics.csv")
```

